Density-dependent selection and the limits of relative fitness

### 2017-12-04 manuscript.tex
l

\title{Density-dependent selection and the population ecology of relative fitness}
%\author{Jason Bertram $^{1,\ast}$ \\ 
%Joanna Masel $^{1}$}

@@

There are a variety of different measures of fitness. Some widely used examples in evolutionary ecology are expected lifetime reproductive ratio $R_0$, intrinsic growth rate $r$, saturation population density (often labeled ``$K$'') \citep{benton_2000}, and invasion fitness \citep{metz_1992}. In addition, ``relative fitness'' is the standard in much of evolutionary biology, particularly evolutionary genetics, where attention is generally restricted to relative genotypic proportions \cite[pp. 468]{barton_2007}. This variety is not necessarily problematic in itself, because different measures of fitness may be more useful in different circumstances. But any measure of fitness should ultimately be grounded in the processes of birth and death which govern population biology \citep{metcalf_2007,doebeli_2017}. While this grounding is clear for absolute fitness measures like $r$ and $K$, relative fitness seems largely divorced from population ecology.

In uncrowded populations, relative fitness simply represents differences in the intrinsic exponential growth rate $r$ \citep[pp. 26]{crow_1970}, with selection favoring greater $r$ ($r$-selection). In crowded populations, relative fitness models almost universally assume that total population size $N$ is fixed, or has some externally imposed time course. This is exemplified by the Wright-Fisher model, in which time advances in discrete non-overlapping generations, and fitness can be interpreted as a product of fertility and juvenile viability \citep[pp. 185]{crow_1970}. The limitations of these relative fitness models are openly acknowledged, usually with an emphasis on the difficulty of incorporating more of the life cycle ``components'' of fitness (e.g. \cite[pp. 276]{ewens_2004}). But more fundamentally, the constant-$N$, relative fitness description uncouples selection and demography, at odds with the fact that the same birth/death events drive both.

The issue can be expressed more formally by revisiting MacArthur's analysis of selection in crowded populations \citep{macarthur_1967}. MacArthur considers a population consisting of two types with densities $n_1$ and $n_2$ subject to density-dependent population growth described by
\begin{equation}
\frac{d n_1}{d t}=f_1(n_1,n_2)\qquad\frac{d n_2}{d t}=f_2(n_1,n_2). \label{eq:macgeneral}
\end{equation}
Apart from the type densities, the environment is assumed to remain constant. The functions $f_1$ and $f_2$ must decline to zero if $n_1$ or $n_2$ are sufficiently large, because no population has unlimited resources. This defines the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$ in $(n_1,n_2)$ space. The outcome of selection is then determined by the relationship between these nullclines. Specifically, a type will be excluded if its nullcline is completely contained in the region bounded by the other type's nullcline. In other words, for a type to have the possibility of persisting, it must be able to keep growing to higher densities than the other type can tolerate in some region of $(n_1,n_2)$ space (Fig.~\ref{fig:Ksel}a).

The connection between constant-$N$ relative fitness models and the broad class of density-regulated growth models considered by MacArthur (Eq.~\ref{eq:macgeneral}) is not clear, since the former only describe selection along lines defined by $n_1+n_2=N$ for each $N$ (Fig.~\ref{fig:Ksel}b). In particular, the $f_1$ and $f_2$ nullclines are not even defined. It is clear, however, that constant-$N$ relative fitness models are not compatible with some of the most widely-used models of density-regulated population growth: 1) In the logistic model, the type with the greatest saturation population density excludes the others. 2) The ``$R^*$ rule'', a central tenet of resource competition theory, states that when growth is limited by a single homogeneous consumable resource, the type able to deplete the resource to the lowest equilibrium density $R^*$ excludes the others \citep{grover_1997}. Differences in $R^*$ will often entail differences in saturation density. 3) The Lotka-Volterra competition model also couples selection in crowded populations to changes in total density $N$ (we return to this in section ``$K$-selection, $c$-selection and relative fitness''). It would therefore seem that the ubiquitous constant-$N$, relative fitness description of selection is incompatible with a huge class of population ecological processes occurring in nature and experiments.

\begin{figure}
\centering
\includegraphics[scale=0.8]{Kplot.pdf}
\caption{\label{fig:Ksel} (a) MacArthur's dynamical argument for how selection operates in crowded environments, illustrated using the logistic model $f_1=r_1(1-\frac{n_1+n_2}{K_1})n_1$ and $f_2=r_2(1-\frac{n_1+n_2}{K_2})n_1$ in Eq. \eqref{eq:macgeneral}. Here $r_1=r_2$ and $K_1>K_2$. (b) The constant-$N$, relative fitness description of selection.}
\end{figure}

The constant-$N$, relative fitness description has historically been justified as a short-term linear approximation \cite[pp. 277]{ewens_2004}. That is, within a sufficiently short time frame, population size can be treated as constant and selection quantified in terms of constant selection coefficients expressing relative fitness differences. Provided that selection is sufficiently weak and stable over time, this ``short-term''  assumption is not a major restriction. Yet it is increasingly recognized that selection is not always weak, that it can fluctuate considerably over time, and that $N$ can vary by orders of magnitude over a few generations as a routine feature of a population's ecology. These are not rare exceptions, but occur widely in nature and the lab, including in wild \textit{Drosophila} \citep{messer_2016}. The short-term approximation also precludes consideration of inherently long-term evolutionary processes like the management of genetic load and population extinction \citep{bertram_2017}. Nevertheless, relative fitness models like Wright-Fisher are the foundation for much of the population genetic literature, and are still widely used without considering the ``short-term'' restriction. Thus, to summarize, there are important theoretical and practical reasons to deepen our understanding of the population-ecological foundations of relative fitness models.

@@

Each iteration, adults produce new offspring (``propagules''), $m_i$ of which disperse to unoccupied territories. We assume that adults cannot be ousted from their territories, so that $m_i$ only includes propagules landing on unoccupied territories. Propagules disperse at random over the unoccupied territories, regardless of distance from their parents, and independently of each other. There is no interaction between propagules (e.g. avoidance of territories crowded with propagules). Loss of propagules during dispersal is subsumed into $m_i$. We assume that each adult produces a constant number $b_i$ of successfully dispersing propagules; the loss of propagules due to dispersal to occupied territories then implies $m_i=b_i(1-N/T)n_i$. Note that due to our assumption of uniform dispersal, the parameter $b_i$ can be thought of as a measure of ``colonization ability'', which combines fecundity and dispersal ability \citep{levins_71,tilman_94,bolker_99}. In addition to random dispersal, we will compare our model to perfect directed dispersal, in which each propagule finds an unoccupied territory if one is available ($m_i=b_i$) \citep{chesson_1983}.

@@

%type $i$ can win at most $m_i$ territories, yet Eq.~\eqref{eq:lottery} demands $c_i l_i/\sum_j c_j l_j$ of the $U$ unoccupied territories, for any value of $U$.

@@

\subsection*{Mean Field Approximation}

@@

\subsection*{$K$-selection, $c$-selection and relative fitness}

We now compare the density-dependent lottery model from the previous section with MacArthur's claims about selection in crowded environments.

As shown in the Introduction, MacArthur's argument revolves around the behaviour of  ecological models of the general form Eq.~\ref{eq:macgeneral}, behaviour which depends on the relationship between the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$. To formalize this relationship, MacArthur used the symbol ``$K$'' to label the four intersection points of the nullclines with the $n_1$ and $n_2$ axes, specifically $f_1(K_{11},0)=0$, $f_1(0,K_{12})=0$, $f_2(0,K_{22})=0$ and $f_2(K_{21},0)=0$. These $K$ values determine whether a region of higher-density growth exists for each type, provided that the nullclines are close to being straight lines. Note that only $K_{11}$ and $K_{22}$ are saturation densities akin to the $K$ parameter in the logistic model; following widespread convention, we will refer to selection on these saturation densities as ``$K$-selection'' (Fig.~\ref{fig:Ksel}a). The other intersection points, $K_{12}$ and $K_{21}$, are related to competition between types. For instance, in the Lotka-Volterra competition model we have
\begin{align}
\frac{d n_1}{dt} = r_1(1-\alpha_{11}n_1-\alpha_{12}n_2)n_1 = f_1(n_1,n_2)\nonumber\\
\frac{d n_2}{dt} = r_2(1-\alpha_{22}n_1-\alpha_{21}n_2)n_2 = f_2(n_1,n_2)\label{eq:LV}
\end{align}
where $\alpha_{11}=1/K_{11}$ and $\alpha_{22}=1/K_{22}$ measure competitive effects within each type, while $\alpha_{12}=1/K_{12}$ and $\alpha_{21}=1/K_{21}$ measure competitive effects on the first type due to the second (Fig.~\ref{fig:LVvslottery}a).

Thus, when MacArthur concludes that  ``fitness is $K$'' in crowded populations \citep[pp. 149]{macarthur_1967}, it does not imply a dichotomy between $r$-selection (uncrowded) and $K$-selection (crowded) in the sense of selection for greater saturation density. MacArthur's argument shows that selection either favors the ability to keep growing at ever higher densities (moving a type's own nullcline outwards), or the ability to suppress the growth of competitors at lower densities (moving the nullcline of competitors inwards) \citep{gill_1974}. This general idea applies even if the nullclines are nonlinear to such an extent that the ``$K$'' values themselves do not give much information about the regions of high-density growth.

@@

Constant-$N$ relative fitness models are not compatible with $K$-selection, and must refer to the competitive effects summarized by $K_{12}$ and $K_{21}$. We thus now restrict our attention to the case $K_{11}=K_{22}$. Fig.~\ref{fig:LVvslottery} shows two examples of selection between two types with the same saturation density. Fig.~\ref{fig:LVvslottery}a shows exclusion by a superior Lotka-Volterra competitor ($\alpha_{21}>\alpha_{12}$). Even though the saturation densities are identical, the density trajectories do not follow a line of constant $N$. This is a general properly of the Lotka-Volterra competition model; only a highly restricted subset of $r$ and $\alpha$ values will keep $N$ constant over a selective sweep (further details in Appendix C). Intuitively, for one type to exclude another with the same saturation density, inter-type competitive effects must be stronger than intra-type competitive effects, causing a dip in $N$ over a selective sweep.

By contrast, if one type in our density-dependent lottery model has a $c$ advantage but birth and death rates are identical, the density trajectories converge on the line of constant density equal to the saturation density. Selection then occurs purely along this line, uncoupled from the density regulation of $N$. In other words, once the population reaches demographic equilibrium, it behaves indistinguishably from a constant-$N$ relative fitness model. More generally, the competitive ability trait $c$ does not directly affect population density (this can be seen formally in Eq.~\eqref{eq:Nmaster}), since $c$ only affects which type wins a territory, not whether a territory is won at all. In other words, $c$ is a zero-sum relative fitness trait embedded within a non-zero-sum fitness model.

Thus, we can conclude that relative fitness models are perfectly consistent with MacArthur's claim that ``fitness is $K$'' in crowded populations, bearing in mind his broad use of the symbol ``$K$''. However, inconsistency remains with standard growth models such as the Lotka-Volterra model. We revisit this issue in the Discussion.

%, we will evaluate the broader ecological plausibility of relative fitness models from the basis of our density-dependent lottery model.

@@

If s is small it doesn't matter anyway

%The strength of this assumption is clear if we note that essence of the dichotomy underlying the $r$/$K$ scheme is between interaction-dependent selection and interaction-independent selection. That is, selective shifts in frequency are a result of differences in absolute growth rates, but these differences can arise in two logically distinct ways: 1) some types expand more rapidly the absence of interactions between individuals or 2) some types are superior in their interactions with other types. Population density is a key factor controlling whether individuals interact, thereby setting the relative contributions of these forms of selection. But there are obviously important forms of interaction that are not expressible in terms of slanted density isoclines. For instance, contests for territory and mates are predominantly relative. MacArthur's argument starts with assumptions about how selection depends on density, and then tries to make sense of what this says about the evolution of competitive traits such as ``efficiency'' of consumable resource consumption. The study of density-dependent selection should start from a description of the interactions between types, and then analyze how the effects of those interactions vary with density.

>>>>

\title{Density-dependent selection and the limits of relative fitness}
%\author{Jason Bertram $^{1,\ast}$ \\ 
%Joanna Masel $^{1}$}

@@

There are a variety of different measures of fitness. Some widely used examples in evolutionary ecology are expected lifetime reproductive ratio $R_0$, intrinsic growth rate $r$, saturation population density (often labeled ``$K$'') \citep{benton_2000}, and invasion fitness \citep{metz_1992}. In addition, ``relative fitness'' is the standard in much of evolutionary biology, particularly evolutionary genetics, where attention is generally restricted to relative genotypic proportions \cite[pp. 468]{barton_2007}. The variety of fitness measures is not problematic in itself, because different measures  may be more useful in different circumstances. But it should be clear how the measure being used is connected to the processes of birth and death which govern population biology \citep{metcalf_2007,doebeli_2017}. While such a connection is fairly clear for absolute fitness measures like $r$, relative fitness seems largely divorced from population biology. It has even been proposed that relative fitness be justified from more abstract measure-theoretical arguments, abandoning population biology altogether \citep{wagner_2010}.

In uncrowded populations, relative fitness simply represents differences in intrinsic growth rate. In continuous time, we have the canonical selection equation \citep[pp. 26]{crow_1970}
\begin{equation}
\frac{d p_i}{dt}=(r_i-\overline{r})p_i, \label{eq:canonical}
\end{equation}
where $r_i$ is the intrinsic exponential growth rate of type $i$, $p_i$ is its frequency, and $\overline{r}=\sum_i r_i p_i$ is the population mean $r$. If there are two types present, a wildtype and a mutant for instance, then Eq.~\eqref{eq:canonical} can be written as $\frac{d p_i}{dt}=s p_i(1-p_i)$, where the constant selection coefficient $s$ is the difference in $r$ between types; the corresponding adaptive sweeps follow a logistic curve.

The difficulty with Eq.~\eqref{eq:canonical} arises in crowded populations. It is then commonly assumed that total population density $N$ has reached its equilibrium value, which is assumed to be a fixed constant. This is exemplified by the Wright-Fisher model. Yet MacArthur famously showed that when population growth is density-regulated, selection in crowded populations is intimately connected to the ability to keep growing at higher densities than other types can tolerate \citep{macarthur_1967}. The classic example is the logistic model, where the type with the greatest saturation population density ``$K$'' excludes the others. Similarly, the ``$R^*$ rule'', a central tenet of resource competition theory, states that when growth is limited by a single homogeneous consumable resource, the type able to deplete the resource to the lowest equilibrium density $R^*$ excludes the others \citep{grover_1997}. Differences in $R^*$ will often entail differences in saturation density. The Lotka-Volterra competition model also couples selection in crowded populations to changes in $N$ except in a few special cases (we return to this in section ``$K$-selection, $c$-selection and relative fitness''). Thus, $N$ changes during, and as a result of, the adaptive sweeps described by Eq.~\eqref{eq:canonical}. It would therefore seem that the ubiquitous constant-$N$, relative fitness description of selection is incompatible with a huge class of population ecological processes occurring in nature and experiments.

\begin{figure}
\centering
\includegraphics[scale=0.8]{Kplot.pdf}
\caption{\label{fig:Ksel} Selection in crowded environments shown as a phase diagram for the densities of two types $n_1$ and $n_2$. (a) The logistic model $\frac{dn_1}{dt}=r_1(1-\frac{n_1+n_2}{K_1})n_1$ and $\frac{dn_2}{dt}=r_2(1-\frac{n_1+n_2}{K_2})n_1$ with $r_1=r_2$ and $K_1>K_2$. (b) The constant-$N$, relative fitness description of selection.}
\end{figure}

The relative fitness description has been justified in broadly two different ways for crowded populations (we do not address Wagner's [\citeyear{wagner_2010}] measure theorical justification, which explicitly seeks to be independent of population biology and thus falls outside of our scope). First, if we only care about type frequencies, we can sometimes simply ignore $N$ \cite[pp. 468]{barton_2007}, because Eq.~\eqref{eq:canonical} is unaffected by changes in $N$ if the mechanism regulating density applies equally to all types \citep{prout_1980}. Yet, while this does allow us to relax the restrictive assumption of constant $N$, it does not address the assumption of constant $s$. In the examples from the previous paragraph, different types \textit{are} affected by the regulation of density in different ways; indeed, the type-specific responses to density  are at the center of MacArthur's argument and the literature on density-dependent selection that grew out of it.

Second, constant $N$ and $s$ can both be seen as a short-term linear approximation \cite[pp. 277]{ewens_2004}. That is, within a sufficiently short time frame, $N$ and $s$ can be treated as constant. Provided that selection is sufficiently weak and stable over time, this ``short-term''  assumption is not a major restriction \cite[PAGE NUMBER]{crow_1970}. Yet it is increasingly recognized that selection is not always weak, that it can fluctuate considerably over time \citep{messer_2016}, and that $N$ can vary by orders of magnitude over a few generations as a routine feature of a population's ecology These are not rare exceptions, but occur widely in nature and the lab, including in wild \textit{Drosophila} \citep{bergland_14}. Nevertheless, relative fitness models like Wright-Fisher are the foundation for much of the population genetic literature, and are still widely used without considering the ``short-term'' restriction or the gulf with population ecology \citep{mallet_2012}. Thus, it is important that we deepen our  understanding of the population ecological basis of relative fitness models, both to gain insight into their domain of applicability, and as part of the broader challenge of synthesizing ecology and evolution.

%The constant-$N$ relative fitness description of selection also precludes consideration of inherently long-term evolutionary processes like the management of genetic load and population extinction \citep{bertram_2017}. Adaptive dynamics provides a powerful framework for modeling the interplay between long-term evolution and population ecology, including complex couplings between evolutionary change and population density. However, the focus of this latter work is on trait evolution rather than the underlying genetics, and in particular, selective sweeps are typically subsumed into effectively-instantaneous ``trait substitutions''. We emphasize that our focus here is the description of selection itself, which particularly critical for making sense of evolution at the genetic level.

%The limitations of relative fitness models are openly acknowledged, usually with an emphasis on the difficulty of incorporating multiple life cycle ``components'' of fitness (e.g. \cite[pp. 276]{ewens_2004} \citep{prout_1980}).

%in which time advances in discrete non-overlapping generations and fitness is interpreted as a product of fertility and juvenile viability \citep[pp. 185]{crow_1970}.

%But more fundamentally, the constant-$N$, relative fitness description uncouples selection and demography, at odds with the fact that the same birth/death events drive both.

 %The connection between constant-$N$ relative fitness models and the broad class of density-regulated growth models considered by MacArthur (Eq.~\ref{eq:macgeneral}) is not clear, since the former only describe selection along lines defined by $n_1+n_2=N$ for each $N$ (Fig.~\ref{fig:Ksel}b). In particular, the $f_1$ and $f_2$ nullclines are not even defined.

@@

Each iteration, adults produce new offspring (``propagules''), $m_i$ of which disperse to unoccupied territories. We assume that adults cannot be ousted from their territories, so that $m_i$ only includes propagules landing on unoccupied territories. Propagules disperse at random over the unoccupied territories, regardless of distance from their parents, and independently of each other. There is no interaction between propagules (e.g. avoidance of territories crowded with propagules). Loss of propagules during dispersal is subsumed into $m_i$. We assume that each adult produces a constant number $b_i$ of successfully dispersing propagules; the loss of propagules due to dispersal to occupied territories then implies $m_i=b_i(1-N/T)n_i$. Note that due to our assumption of uniform dispersal, the parameter $b_i$ can be thought of as a measure of ``colonization ability'', which combines fecundity and dispersal ability \citep{levins_71,tilman_94,bolker_99}.

@@

@@

\subsection*{Mean-field approximation of the density-dependent lottery}

@@

\subsection*{$K$-selection versus relative fitness}

We now compare the density-dependent lottery model from the previous section with MacArthur's analysis of selection in crowded environments \citep{macarthur_1967}. MacArthur considers a population with two types that have densities $n_1$ and $n_2$ subject to density-dependent growth described by
\begin{equation}
\frac{d n_1}{d t}=f_1(n_1,n_2)\qquad\frac{d n_2}{d t}=f_2(n_1,n_2). \label{eq:macgeneral}
\end{equation}
The environment is assumed to remain constant apart from the type densities. The functions $f_1$ and $f_2$ must decline to zero if $n_1$ or $n_2$ are sufficiently large, because no population has unlimited resources. This defines the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$ in $(n_1,n_2)$ space. The outcome of selection is then determined by the relationship between these nullclines. Specifically, a type will be excluded if its nullcline is completely contained in the region bounded by the other type's nullcline. In other words, for a type to have the possibility of persisting, it must be able to keep growing to higher densities than the other type can tolerate in some region of $(n_1,n_2)$ space (Fig.~\ref{fig:Ksel}a).

To formalize the relationship between nullclines, MacArthur used the symbol ``$K$'' to label the four intersection points of the nullclines with the $n_1$ and $n_2$ axes, specifically $f_1(K_{11},0)=0$, $f_1(0,K_{12})=0$, $f_2(0,K_{22})=0$ and $f_2(K_{21},0)=0$. These $K$ values determine whether a region of higher-density growth exists for each type, provided that the nullclines are close to being straight lines. Note that only $K_{11}$ and $K_{22}$ are saturation densities akin to the $K$ parameter in the logistic model; following widespread convention, we will refer to selection on these saturation densities as ``$K$-selection'' (Fig.~\ref{fig:Ksel}a). The other intersection points, $K_{12}$ and $K_{21}$, are related to competition between types. For instance, in the Lotka-Volterra competition model we have
\begin{align}
f_1(n_1,n_2) = r_1(1-\alpha_{11}n_1-\alpha_{12}n_2)n_1\nonumber\\
f_2(n_1,n_2) = r_2(1-\alpha_{22}n_1-\alpha_{21}n_2)n_2\label{eq:LV}
\end{align}
where $\alpha_{11}=1/K_{11}$ and $\alpha_{22}=1/K_{22}$ measure competitive effects within each type, while $\alpha_{12}=1/K_{12}$ and $\alpha_{21}=1/K_{21}$ measure competitive effects on the first type due to the second (Fig.~\ref{fig:LVvslottery}a).

Thus, when MacArthur concludes that  ``fitness is $K$'' in crowded populations \citep[pp. 149]{macarthur_1967}, the meaning is that selection either favors the ability to keep growing at ever higher densities (moving a type's own nullcline outwards), or the ability to suppress the growth of competitors at lower densities (moving the nullcline of competitors inwards) \citep{gill_1974}. This general idea applies even if the nullclines are nonlinear to such an extent that the ``$K$'' values themselves do not give much information about the regions of high-density growth.

@@

It is obvious from Eq.~\eqref{eq:LV} that selection can favor a superior competitor in a crowded population even if its saturation density is the same as, or lower than that of the other types present. However, note that the Lotka-Volterra model still closely couples selection to population density. Fig.~\ref{fig:LVvslottery}a shows Lotka-Volterra selection between two types with the same saturation density ($\alpha_{11}>\alpha_{22}$, $\alpha_{21}>\alpha_{12}$). Even though the initial and final densities of a sweep are the same, density is not constant during a sweep. Only a highly restricted subset of $r$ and $\alpha$ values will keep $N$ constant over a selective sweep (further details in Appendix C). Intuitively, for one type to exclude another with the same saturation density, inter-type competitive effects must be stronger than intra-type competitive effects, causing a dip in $N$ over the sweep.

By contrast, if one type in our density-dependent lottery model has a $c$ advantage but birth and death rates are identical, the density trajectories converge on the line of constant density equal to the saturation density Fig.~\ref{fig:LVvslottery}b. Selection then occurs purely along this line, uncoupled from the density regulation of $N$. In other words, once the population reaches demographic equilibrium, it behaves indistinguishably from a constant-$N$ relative fitness model. More generally, the competitive ability trait $c$ does not directly affect population density (this can be seen formally in Eq.~\eqref{eq:Nmaster}), since $c$ only affects which type wins a territory, not whether a territory is won at all. This is all perfectly consistent with MacArthur's general argument.

\section*{The limits of relative fitness}

We are now in a position to analyze the validity of relative fitness models more rigorously. To do so, we evaluate some objections to Eq.~\eqref{eq:canonical} in a crowded poulation using our density-dependent lottery model as a benchmark to better understand the population ecological implications.

We first set aside the most obvious objection: the selection coefficient will often be density-dependent because crowded and uncrowded conditions are so different. This objection cannot be ignored if $N$ changes substantially during a sweep. We return to this in the Discussion. For now we restrict our attention to the case that $N$ is at its equilibrium value.

With $N$ at equilibrium, the presence of density-dependent selection is beside the point. As we have seen in our lottery model, the importance of $c$ depends on density, yet since $c$ has no effect on density, $s$ is exactly constant for $c$ sweeps. For $b$ and $d$ sweeps, $s$ is also exactly consant, because the overall regulation of population density applies equally to all types \citep{prout_1980}. To see this formally, we set $c_i$ to be the same for all types in Eq.~\eqref{eq:master} which yields
\begin{equation}
\Delta n_i = \frac{m_i}{M}(1-e^{-L})U-d_i n_i = \left(b_i\frac{1-e^{-L}}{L}-d_i \right) n_i.
\end{equation}
Thus, the density-regulation of population size occurs via the juvenile recruitment fraction $(1-e^{L})/L$. This applies uniformly to all types and therefore does not induce density-dependence in $b$ or $d$ selective advantages. Thus, $b$, $c$ and $d$ sweeps all obey Eq.~\eqref{eq:canonical} exactly. That is, in our lottery model density will adjust itself as a result of selection, selection is density-dependent, and yet Eq.~\eqref{eq:canonical} applies exactly for pure $b$, $c$ and $d$ sweeps.

As noted in the Introduction, widely used models of population growth predict that selection in crowded populations will intrinsically affect population density. In the previous section we revisited MacArthur's argument and saw that there is no reason to expect this to be true in general, and indeed the competitive ability trait $c$ in our lottery model does not have this property.

@@

It is important to recognize that MacArthur's argument does not imply a dichotomy between $r$-selection (uncrowded) and $K$-selection (crowded), with the latter taken to mean selection for greater saturation density. A better dichotomy is between interaction-dependent selection and interaction-independent selection. That is, selective shifts in frequency are a result of differences in absolute growth rates, but these differences can arise in two logically distinct ways: 1) some types expand more rapidly in the absence of interactions between individuals or 2) some types are superior in their interactions with other types. Population density is a key factor controlling whether individuals interact, thereby setting the relative contributions of these forms of selection.
